---
title: 'Session 8: Function Writing'
author: "Samuel P Callisto"
date: "August 2, 2018"
output: pdf_document
---

\tableofcontents
\newpage

```{r, include=F, warning=F, message=F}
## install packages if necessary
packages <- c("testthat","tidyverse")
if(any(!(packages %in% installed.packages()[,"Package"]))){
  install.packages(packages[which(!packages %in% installed.packages()[,"Package"])])
}

## load packages
library(testthat)
library(tidyverse)
```


## Function Writing

### A trivial example
```{r}
sumMinusOne <- function(x){
  output <- 0
  for(i in 1:length(x)){
    output <- output + x[i]
  }
  return(output-1)
}

sumMinusOne(2:4)
```

## When would I ever use this?

### Example 1: Wrapper function
Problem: importing files with multiple headers causes the data type to be interpretted incorrectly, causing resulting in manual typecasting for multiple rows (annoying!)
```{r, warning=F}
## Excel files
excel <- read.csv("datasets/TPM_sim_dataset_20180607.csv", as.is = T, header = T)

## dataset imported using read.csv()
str(excel)
excel$subjectid <- as.integer(excel$subjectid)
excel$DV <- as.double(excel$DV)
```

Solution: write a wrapper function that assigns correct header row while maintaining data types.
```{r}
headr <- function(file, header.row=1, data.start=3){
  headers <- read.csv(file = file, skip=header.row-1, header = F, nrows = 1, as.is = T)
  dataset <- read.csv(file=file, skip = data.start-1, header = F, as.is=T)
  names(dataset) <- headers
  return(dataset)
}

## import same file using headr wrapper function
topiramateData <- headr("datasets/TPM_sim_dataset_20180607.csv")

## dataset imported using headr()
str(topiramateData)
```
Notice how few arguments need to be filled out manually each time you import a file using the helper function since you are allowed to set your defaults.

Slightly more advanced solution:
You can use the ellipsis operator (...) to pass additional commands into the functions called by your wrapper function. In this example, by adding this to headr(), we can access arguments in read.csv()
```{r}
headr <- function(file, header.row=1, data.start=3, ...){
  headers <- read.csv(file = file, skip=header.row-1, header = F, nrows = 1, as.is = T)
  dataset <- read.csv(file=file, skip = data.start-1, header = F, as.is=T, ...)
  names(dataset) <- headers
  return(dataset)
}

## import same file using headr wrapper function
topiramateData <- headr("datasets/TPM_sim_dataset_20180607.csv", na.strings=-99)

## dataset imported using headr()
str(topiramateData)
```

### Anonymous Functions
R allows you to call a function without naming it, called an Anonymous Function. This is useful typically used when you are applying a small function to a matrix using apply().
```{r}
apply(topiramateData, 2, range)
```
Due to the missing value in the DV column, we cannot get range values for this column. We need to use an anonymous function to pass the na.rm argument into the range function if we wish to apply it to the entire data.frame.
```{r}
## this approach will only give us an error:
## apply(topiramateData, 2, range(na.rm=T))

## instead we can use an anonymous function to access arguments
apply(topiramateData, 2, function(x) range(x,na.rm=T))
```
By creating an anonymous function, we can get results for all the columns in the data.frame. We can create any sort of anonymous function to use in this context; let's look at a more complicated example.

### Example 2: Dealing with Times
Problem: RedCap stores my times as a character string ("6:45"), but I want to calculate the difference between observations.

Solution: write a function to use this time and the next time you encounter clock times
```{r}
numericTime <- function(vec){
  ## separate hours and minutes into a vector
  sapply(strsplit(vec,":"),
         function(x) {
           ## convert to numeric type to allow arithmetic operations
           x <- as.numeric(x)
           ## numeric time = hours + (minutes/60) rounded to two decimals
           round(x[1]+x[2]/60,2)
         }
  )
}

topiramateData$DECTIME <- numericTime(topiramateData$TIME)
```
\underline{Challenge:} Use RStudio's built-in debug functions to verify that x <- as.numeric(x) is necessary in our function. 


## A general rule of thumb:
If you find yourself copying code within or between files, you should probably just write a function. Better still, add functions to a personal R package that you can easily import and share.

### Great resource for learning more about writing functions
http://adv-r.had.co.nz/Functional-programming.html

\newpage

## Unit testing
How can you check if your function is doing what you think it is? Let's go back to our sumMinusOne() function

### A trivial example revisited
```{r}
test_that("single value minus one",{
  expect_equal(1, sumMinusOne(2))
})

test_that("vector sum minus one",{
  expect_equal(5, sumMinusOne(1:3))
})

test_that("characters shouldn't work",{
  expect_error(sumMinusOne("test"))
  expect_error(sumMinusOne("1"))
})
```

There is no message if the test passes, but you will get an error if the tests fail. That means you can write multiple tests into your code and let them be tested automatically, alerting you if the change you just made altered the functionality in an unexpected way.

```{r}
try(
test_that("characters shouldn't work",{
  expect_error(sumMinusOne(1))
})
, outFile = stdout())
```

### Testing our wrapper function for read.csv()
```{r}
test_that("ID imported as integer",{
  imported <- headr("datasets/TPM_sim_dataset_20180607.csv")
  expect_true(is.integer(imported[1,1]))
})

test_that("time imported as character",{
  imported <- headr("datasets/TPM_sim_dataset_20180607.csv")
  expect_true(is.character(imported[1,3]))
})

test_that("concentrations imported as numeric",{
  imported <- headr("datasets/TPM_sim_dataset_20180607.csv")
  expect_true(is.numeric(imported[1,4]))
})
```

### Testing numericTime()
```{r}
test_that("numbers shouldn't work",{
  expect_error(numericTime(5.37))
})

test_that("strings without colon shouldn't work",{
  expect_true(is.na(numericTime("546")))
  expect_true(is.na(numericTime("5.46")))
})

test_that("single digit & double digit times should work",{
  expect_equal(5.5, numericTime("5:30"))
  expect_equal(12.0, numericTime("12:00"))
  ## notice we don't test for real clock times
  ## we could add this into the function later
  expect_equal(25.75, numericTime("25:45"))
})

test_that("differences in time can now be calculated", {
  expect_equal(2, numericTime("5:00") - numericTime("3:00"))
  expect_equal(4.5, numericTime("4:30"), numericTime("12:00"))
})

test_that("vectorized application succeeds",{
  expect_equal(c(12, 1.5), numericTime(c("12:00", "1:30")))
})
```

