---
title: 'Session 7: Function Writing'
author: "Samuel P Callisto"
date: "June 9, 2018"
output: pdf_document
---

```{r, include=F}
## install packages if necessary
if(!(c("testthat","tidyverse") %in% installed.packages()[,"Package"])) install.packages(c("testthat", "tidyverse"))
```
```{r, warning=F, message=F}
## load packages
library(testthat)
library(tidyverse)
```


## Function Writing

### A trivial example
```{r}
sumMinusOne <- function(x){
  output <- 0
  for(i in 1:length(x)){
    output <- output + x[i]
  }
  return(output-1)
}

sumMinusOne(2:4)
```

## When would I ever use this?

### Example 1: Wrapper function
Problem: importing files with multiple headers causes the data type to be interpretted incorrectly, causing resulting in manual typecasting for multiple rows (annoying!)
```{r, warning=F}
## Excel files
excel <- read.csv("datasets/TPM_sim_dataset_20180607.csv", as.is = T, stringsAsFactors = T, header = T)

## dataset imported using read.csv()
str(excel)
excel$subjectid <- as.integer(excel$subjectid)
excel$var3 <- as.double(excel$var3)
```
Solution: write a wrapper function that assigns correct header row while maintaining data types.
```{r}
headr <- function(file, header.row=1, data.start=3){
  headers <- read.csv(file = file, skip=header.row-1, header = F, nrows = 1, as.is = T)
  dataset <- read.csv(file=file, skip = data.start-1, header = F, as.is=T)
  names(dataset) <- headers
  return(dataset)
}

## import same file using headr wrapper function
topiramateData <- headr("datasets/TPM_sim_dataset_20180607.csv")

## dataset imported using headr()
str(topiramateData)
```
Notice how few arguments need to be filled out manually each time you import a file using the helper function since you are allowed to set your defaults.

### Example 2: Dealing with Times
Problem: RedCap stores my times as a character string ("6:45"), but I want to calculate the difference between observations.

Solution: write a function to use this time and the next time you encounter clock times
```{r}
numericTime <- function(vec){
  ## separate hours and minutes into a vector
  sapply(strsplit(vec,":"),
         function(x) {
           ## convert to numeric type to allow arithmetic operations
           x <- as.numeric(x)
           ## numeric time = hours + (minutes/60) rounded to two decimals
           round(x[1]+x[2]/60,2)
         }
  )
}

topiramateData$TIME <- numericTime(topiramateData$var2)
```

## A general rule of thumb:
If you find yourself copying code within or between files, you should probably just write a function. Better still, add functions to a personal R package that you can easily import and share.

### Great resource for learning more about writing functions
http://adv-r.had.co.nz/Functional-programming.html

## Unit testing
How can you check if your function is doing what you think it is?

\newpage